# Using `ggplot2`

## Overview

`ggplot2` is the primary visualization package in R. It is based on the **grammar of graphics**, a system for building plots by combining components.

```r
ggplot(data, aes(...)) +
  geom_*()
```

- `data`: dataset  
- `aes()`: maps variables to visual properties  
- `geom_*()`: defines how data are drawn  
- `+`: adds layers  

## Basic Scatterplot

```{r}
library(ggplot2)

ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point()
```

## Encode Additional Information

```{r}
ggplot(mtcars, aes(wt, mpg, color = factor(cyl))) +
  geom_point()
```

## Aesthetic Mapping vs Fixed Values

::: {.callout-note}
### Key Rule

Map variables inside `aes()`. Set constants outside `aes()`.
:::

```{r}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point(color = "blue")
```

## Add Layers

```{r}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

## Color Best Practices

::: {.callout-tip}
### Using Color Effectively

- Use color to highlight important differences
- Avoid overly bright or distracting palettes
- Ensure sufficient contrast with the background
- Consider color blindness (use accessible palettes)
:::

```{r}
# Better color palette
ggplot(mtcars, aes(wt, mpg, color = factor(cyl))) +
  geom_point(size = 2) +
  scale_color_brewer(palette = "Set1")
```

```{r}
# Continuous color scale
ggplot(mtcars, aes(wt, mpg, color = mpg)) +
  geom_point(size = 2) +
  scale_color_viridis_c()
```

## Improve Labels and Readability

```{r}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Fuel Efficiency vs Weight",
    x = "Weight",
    y = "Miles per Gallon"
  )
```

## Step 7: Themes

Themes control the overall appearance of your plot.

```{r}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  theme_minimal()
```

```{r}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  theme_classic()
```

### Customizing Themes

```{r}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.title = element_text(size = 12)
  )
```

## Step 8: Compare Groups with Faceting

```{r}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  facet_wrap(~ cyl)
```

## Common Geometries

```{r}
ggplot(mtcars, aes(wt, mpg)) + geom_point()
ggplot(mtcars, aes(factor(cyl))) + geom_bar()
ggplot(mtcars, aes(mpg)) + geom_histogram(bins = 10)
ggplot(mtcars, aes(factor(cyl), mpg)) + geom_boxplot()
```

## Workflow

::: {.callout-tip}
### Recommended Approach

1. Start simple  
2. Confirm it answers your question  
3. Add layers  
4. Improve clarity and aesthetics  
:::



## Summary

- Build plots with `ggplot(data, aes(...))`  
- Add layers with `+`  
- Map variables inside `aes()`  
- Use color intentionally  
- Use themes to improve readability  



## Exercise

Starting with the following plot, do the following:

- Color by `am`  
- Add a smooth line  
- Apply a different theme  
- Use a color palette

```{r}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point()
```


