# Mapping with `leaflet`

## Overview

`leaflet` is an R package for building interactive maps. It is especially useful when your data include locations, coordinates, regions, or boundaries and you want viewers to explore the map directly.

```r
leaflet(data = ...) %>%
  addTiles() %>%
  add*()
```

- `data`: dataset  
- `leaflet()`: initializes the map  
- `addTiles()`: adds a basemap  
- `add*()`: adds markers, circles, polygons, or other layers  
- `%>%`: builds the map step by step  

---

## Basic Map

```{r}
library(leaflet)

leaflet() %>%
  addTiles()
```

---

## Add a Single Marker

```{r}
leaflet() %>%
  addTiles() %>%
  addMarkers(
    lng = -96.7970,
    lat = 32.7767,
    popup = "Dallas, Texas"
  )
```

---

## Map Data from a Dataset

```{r}
leaflet(quakes) %>%
  addTiles() %>%
  addMarkers(
    lng = ~long,
    lat = ~lat
  )
```

---

## Markers vs Circle Markers

::: {.callout-note}
### Key Distinction

- `addMarkers()` places standard map pins
- `addCircleMarkers()` draws scalable points on the map

For most data visualization tasks, `addCircleMarkers()` is preferred.
:::

```{r}
leaflet(quakes) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~long,
    lat = ~lat
  )
```

---

## Encode Size with Magnitude

::: {.callout-note}
### Key Idea

Marker size is one of the most effective ways to encode magnitude in a map.
:::

```{r}
leaflet(quakes) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~long,
    lat = ~lat,
    radius = ~mag
  )
```

### Scaling Marker Size

Raw values are often too small or too large. Scale them for readability:

```{r}
leaflet(quakes) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~long,
    lat = ~lat,
    radius = ~mag * 2,   # simple scaling
    fillOpacity = 0.7
  )
```

You can also normalize:

```{r}
leaflet(quakes) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~long,
    lat = ~lat,
    radius = ~sqrt(mag) * 3,
    fillOpacity = 0.7
  )
```

---

## Add Rich Popups and Labels

```{r}
leaflet(quakes) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~long,
    lat = ~lat,
    radius = ~mag * 2,
    popup = ~paste(
      "Magnitude:", mag,
      "<br>Depth:", depth
    )
  )
```

---

## Use Color Intentionally

::: {.callout-tip}
### Color Best Practices

- Use color to encode a meaningful variable
- Ensure contrast with the basemap
- Avoid too many categories
:::

```{r}
pal <- colorNumeric(
  palette = "viridis",
  domain = quakes$mag
)

leaflet(quakes) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~long,
    lat = ~lat,
    radius = ~mag * 2,
    color = ~pal(mag),
    fillOpacity = 0.7
  )
```

---

## Add a Legend

```{r}
leaflet(quakes) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~long,
    lat = ~lat,
    radius = ~mag * 2,
    color = ~pal(mag),
    fillOpacity = 0.7
  ) %>%
  addLegend(
    "bottomright",
    pal = pal,
    values = ~mag,
    title = "Magnitude"
  )
```

---

## Adjust the View

```{r}
leaflet() %>%
  addTiles() %>%
  setView(lng = -96.7970, lat = 32.7767, zoom = 10)
```

---

## Change the Basemap

```{r}
leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron)
```

---

## Interactivity

::: {.callout-note}
### Why Use `leaflet`?

Users can:
- zoom and pan
- click for details
- explore spatial patterns
:::

---

## Workflow

::: {.callout-tip}
### Recommended Approach

1. Start with a basemap  
2. Add points  
3. Encode magnitude with size  
4. Add color and labels  
5. Refine view and legend  
:::

---

## Summary

- Use `radius` to encode magnitude  
- Scale size for readability  
- Combine size + color for clarity  
- Use popups to avoid clutter  

---

## Exercise

```{r}
leaflet(quakes) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~long,
    lat = ~lat
  )
```

Modify:
- Scale radius by `mag`
- Add color by `mag`
- Add a popup
