# Interactivity with `plotly`

## Overview

`plotly` is a package for building **interactive** visualizations in R. It is especially useful when you want users to hover, zoom, pan, filter visually, and explore the data on their own.

```r
plot_ly(data = ..., x = ..., y = ..., type = ..., mode = ...)
```

- `data`: dataset  
- `x`, `y`: variables mapped to axes  
- `type`: general chart type  
- `mode`: how marks are drawn  
- `%>%`: adds layers and layout options  

## Basic Scatterplot

```{r}
library(plotly)

plot_ly(
  data = mtcars,
  x = ~wt,
  y = ~mpg,
  type = "scatter",
  mode = "markers"
)
```

## Why Use `plotly`?

::: {.callout-note}
### Key Advantage

The main reason to use `plotly` is interactivity.

Users can:
- hover for details
- zoom into regions
- pan across the figure
- toggle traces on and off
- explore patterns without changing the code
:::

## Encode Additional Information

```{r}
plot_ly(
  data = mtcars,
  x = ~wt,
  y = ~mpg,
  color = ~factor(cyl),
  type = "scatter",
  mode = "markers"
)
```

## Mapping vs Fixed Values

::: {.callout-note}
### Key Rule

Map variables with `~variable`. Set constants directly.
:::

```{r}
plot_ly(
  data = mtcars,
  x = ~wt,
  y = ~mpg,
  type = "scatter",
  mode = "markers",
  marker = list(color = "blue")
)
```

## Add Rich Hover Information

```{r}
plot_ly(
  data = mtcars,
  x = ~wt,
  y = ~mpg,
  type = "scatter",
  mode = "markers",
  text = ~paste(
    "Model:", rownames(mtcars),
    "<br>MPG:", mpg,
    "<br>Weight:", wt,
    "<br>Cylinders:", cyl
  ),
  hoverinfo = "text"
)
```

Hover text is one of the most useful features in `plotly`. It lets you keep the figure clean while still giving viewers access to more detail.

## Add Multiple Traces

```{r}
plot_ly(data = mtcars, x = ~wt, y = ~mpg, type = "scatter", mode = "markers") %>%
  add_lines(y = ~fitted(lm(mpg ~ wt, data = mtcars)), name = "Trend line")
```

Interactive plots are often built by adding traces instead of layering geoms.

## Color Best Practices

::: {.callout-tip}
### Using Color Effectively

- Use color to distinguish meaningful groups
- Avoid using too many categories at once
- Keep contrast strong enough for readability
- Use accessible palettes when possible
:::

```{r}
plot_ly(
  data = mtcars,
  x = ~wt,
  y = ~mpg,
  color = ~factor(cyl),
  colors = "Set1",
  type = "scatter",
  mode = "markers"
)
```

```{r}
plot_ly(
  data = mtcars,
  x = ~wt,
  y = ~mpg,
  color = ~mpg,
  colors = "viridis",
  type = "scatter",
  mode = "markers"
)
```

## Improve Labels and Layout

```{r}
plot_ly(
  data = mtcars,
  x = ~wt,
  y = ~mpg,
  type = "scatter",
  mode = "markers"
) %>%
  layout(
    title = "Fuel Efficiency vs Weight",
    xaxis = list(title = "Weight"),
    yaxis = list(title = "Miles per Gallon")
  )
```

## Styling and Theming

In `plotly`, overall appearance is usually controlled through `layout()` rather than a theme system like `ggplot2`.

```{r}
plot_ly(
  data = mtcars,
  x = ~wt,
  y = ~mpg,
  type = "scatter",
  mode = "markers"
) %>%
  layout(
    title = "Fuel Efficiency vs Weight",
    paper_bgcolor = "white",
    plot_bgcolor = "white",
    xaxis = list(showgrid = FALSE),
    yaxis = list(showgrid = FALSE)
  )
```

### Custom Marker Styling

```{r}
plot_ly(
  data = mtcars,
  x = ~wt,
  y = ~mpg,
  type = "scatter",
  mode = "markers",
  marker = list(
    size = 10,
    opacity = 0.7
  )
)
```

## Compare Groups

```{r}
plot_ly(
  data = mtcars,
  x = ~wt,
  y = ~mpg,
  color = ~factor(cyl),
  type = "scatter",
  mode = "markers"
)
```

A common interactive alternative to faceting is to let users isolate groups by clicking the legend.

## Common Chart Types

```{r}
# Scatterplot
plot_ly(mtcars, x = ~wt, y = ~mpg, type = "scatter", mode = "markers")

# Line chart
plot_ly(economics, x = ~date, y = ~unemploy, type = "scatter", mode = "lines")

# Bar chart
plot_ly(mtcars, x = ~factor(cyl), type = "histogram")

# Histogram
plot_ly(mtcars, x = ~mpg, type = "histogram")

# Boxplot
plot_ly(mtcars, x = ~factor(cyl), y = ~mpg, type = "box")
```

## A Brief Note on `ggplotly()`

`plotly` can convert a `ggplot2` graphic into an interactive figure with `ggplotly()`.

```{r}
library(ggplot2)

p <- ggplot(mtcars, aes(wt, mpg)) +
  geom_point()

ggplotly(p)
```

This can be useful for quick conversions, but when interactivity is the goal, it is often better to build the plot directly with `plotly`.

## Workflow

::: {.callout-tip}
### Recommended Approach

1. Start with the simplest interactive chart
2. Decide what users should be able to explore
3. Add hover text, labels, and traces
4. Refine layout for readability
:::

## Summary

- Build interactive plots with `plot_ly()`
- Use hover, zoom, and legend interaction intentionally
- Map variables with `~`
- Use `layout()` to control appearance
- Build directly in `plotly` when interactivity is the main goal

## Exercise

Starting with the following plot, do the following:

- Color by `am`
- Add custom hover text
- Add a trend line
- Improve the layout

```{r}
plot_ly(
  data = mtcars,
  x = ~wt,
  y = ~mpg,
  type = "scatter",
  mode = "markers"
)
```

